---
title: 'R Basics 2'
author: Christopher Gandrud
date: '2021-07-09'
slug: optional-r-basics-2
weight: 3
categories:
  - R
tags:
  - R
  - programming
Categories: []
Description: ''
Tags: [R, functions, if, loops, basics]
---

## 📝 Lesson Preview

-   [Functions](#functions-functions-intro)

-   [Documenting Functions in R](#documenting)

-   Control Flow

    -   [`if` statements](#if)

    -   [`for` loops](#loop-intro)

    -   [`apply()`](#apply-intro)

-   [Debugging and Testing in R](#debugging-intro)

-   [Advanced: Parallel computation](#parallel-intro)

## Functions {#functions-intro}

If objects are the nouns of the R language, functions are the verbs. They do things to objects. Let's use the `mean` function as an example. This function takes the mean of a numeric vector object. Remember our *numeric_vector* object from Lesson 1:

```{r Shownumeric_vector, echo=TRUE}
numeric_vector <- c(2.8, 2, 14.8)

numeric_vector
```

To find the mean of this object, type:

```{r numeric_vector_mean, echo=TRUE}
mean(x = numeric_vector)
```

We use the assignment operator to place a function's output into an object. For example:

```{r numeric_vector_meanAssign, echo=TRUE}
numeric_vector_mean <- mean(x = numeric_vector)
```

Notice that we typed the function's name then enclosed the object name in parentheses immediately afterwards. This is the basic syntax that all functions use, i.e. `FUNCTION(ARGUMENTS)`. Even if you don't want to explicitly include an argument, *you still need to type the parentheses after the function*.

### Arguments

Arguments modify what functions do. In our most recent example, we gave the `mean` function one argument (`x = numeric_vector`) telling it that we wanted to find the mean of *numeric_vector*. Arguments use the `ARGUMENT_LABEL = VALUE` syntax. In this case, **x** is the argument label.

To find all of the arguments that a function can accept, look at the **Arguments** section of the function's help file. To access the help file, type: `?FUNCTION`. For example:

```{r Ch3HelpMean, echo=TRUE, eval=FALSE, tidy=FALSE}
?mean
```

The help file will also tell you the default values that the arguments are set to. You do not need to explicitly set an argument if you want to use its default value.

You do need to be fairly precise with the syntax for your argument's values. Values for logical arguments must be written as `TRUE` or `FALSE`. Arguments that accept character strings require quotation marks.

Let's see how to use multiple arguments with the `round()` function. This function rounds a vector of numbers. We can use the `digits` argument to specify how many decimal places we want the numbers rounded to. To round the object *numeric_vector_mean* to one decimal place, type:

```{r Ch3Round, echo=TRUE}
round(x = numeric_vector_mean, digits = 1)
```

Note that *arguments are always separated by commas*.

Some arguments do not need to be explicitly labeled. For example, we could write:

```{r Ch3ArgeNoLabel, echo=TRUE}
# Find mean of numeric_vector
mean(numeric_vector)
```

R will do its best to figure out what you want and will only give up when it can't. This will generate an error message. However, to avoid any misunderstandings between yourself and R, it is good practice to label your argument values. This will also make your code easier for other people to read, i.e. it will be more reproducible.

You can stack functions inside of arguments. For example, have R find the mean of *numeric_vector* and round it to one decimal place:

```{r Ch3StackedArgs, echo=TRUE}
round(mean(numeric_vector), digits = 1)
```

Stacking functions inside of each other can create code that is difficult to read. Another option that potentially makes more easily understandable code is piping\index{pipe} using the pipe function (`%>%`) that you can access from the *magrittr* or *dplyr* packages. The basic idea behind the pipe function is that the output of one function is set as the first argument of the next. For example, to find the mean of *numeric_vector* and then round it to one decimal place use:

```{r Ch3Pipe, echo=TRUE}
# Load magrittr package
library(magrittr)
# Find mean of numeric_vector and round to 1 decimal place
mean(numeric_vector) %>%
    round(digits = 1)
```

The `%>%` pipe also allows you to pipe objects to arguments other than the first argument by inserting a `.` where you want the object to be piped to. For example:

```{r PipeAlternate}
TRUE %>% mean(1:10, na.rm = .)
```

The value `TRUE` was piped as the argument to `na.rm` with `na.rm = .`.

From R version 4.1.0, there is a native pipe operator `|>`. It works similar to the *magrittr*/*dplyr* pipe, but is not exactly the same.

```{r}
mean(numeric_vector) |>
    round(digits = 1)
```

### Creating functions

Functions are just objects. You can create your own functions with the `function()` function! Why would you create your own functions? If you ever find yourself repeating code with only minor variation, e.g. just changing the inputs, then it is time to create a function.

For example imagine we want to repeatedly find vector means and the round them to one digit. We can create a new function called `mean_rounder()` to do this for us.

```{r create-mean-rounder}
mean_rounder <- function(x) {
  mean(x) %>% 
    round(digits = 1)
}

mean_rounder(numeric_vector)
```

Notice that the contents of the function are enclosed inside of curly brackets `{}`.

When you make a function, remember to document it. You want to be able to use this function over and over. If it isn't well documented, this will be harder.

## Documenting Functions in R {#documenting}

It is important to get in the habit of documenting your functions as you write them. This makes them easier for others (and yourself in the future) to understand.

There are some R documentation conventions to follow.[^3] Function documentation begins with a description of what the function does. Then each function parameter is documented using `@param`. It is not required, but it is good practice to document the expected type of the value for each argument, e.g. `integer`. You can also give examples of the function working with `@examples`. All of the documentation should be commented out with `#'`. E.g.

[^3]: These conventions are established by the [roxygen2](https://cran.r-project.org/web/packages/roxygen2/vignettes/roxygen2.html) package. If you create an R package to contain your functions, you can use roxygen2 to automatically generate package documentation. For more details see [R Packages](https://r-pkgs.org/).

```{r eval=FALSE}
#' Simulate one A/B test outcome from the binomial distribution
#' and return the difference of the means of the two treatment arms.
#' 
#' @param n integer number of observations per treatment arm
#' @param control_prob numeric probability of success in the control group
#' @param rel_effect numeric relative effect of the treatment compared
#' to the control
one_sim <- function(n, control_prob, rel_effect) {
  treat_prob <- control_prob + (control_prob * rel_effect)
    
  cy <- rbinom(n = n, size = 1, prob = control_prob)
  ty <- rbinom(n = n, size = 1, prob = treat_prob)
  
  mean(ty) - mean(cy)
}
```

## Control flow

We can use conditional evaluation (like if statements) and loops to control the flow of our programs.

### If statements {#if}

So we have seen functions that always (try to) do the same thing for any inputs. There are many cases when you want to do different things depending on the type of input.

For example, imagine we want to create a new binary `[0, 1]` variable from a vector. We want a new vector with a 0 if the original value was less than 5 and 1 if it is greater than or equal to 5. We can use `ifelse()`:

```{r binary-ifelse}
numeric_vector

greater_than_5 <- ifelse(numeric_vector >= 5, 
                         yes = 1, 
                         no = 0) 

greater_than_5
```

The way we read this is if the value is greater than or equal to 5 (`yes`) then the new value is `1`, if not (`no`) then it is `0`.

Note that `>=` is one of R's **logical operators**. See the additional material for a table of common [logical operators](#logical-operators) in R.

For more complex if statements, use `if() {. . .} else()`.

### `for` loops {#loop-intro}

Some times you want to do the same thing to each part of an object. We already saw an example with `ifelse()` which ran an if statement on each element of *numeric_vector*.

A general solution to iterating through elements of an object is the `for()` loop function. Here is an example using R's built in **swiss** data set:

```{r swiss-intro}
head(swiss)
```

For each column in the data set, we'll take the mean, round it to 1 digit, and then print the results with R's `cat()`--concatenate--function.

```{r swiss-for}
for (i in 1:ncol(swiss)) {
    swiss[, i] %>%
    mean() %>%
    round(digits = 1) %>%
    paste(names(swiss)[i], ., "\n") %>%  # the . directs the pipe
    cat()
}


# To do over lunch
means_df <- data.frame()
swiss_names <- names(swiss)

for (i in 1:ncol(swiss)) {
  tmp <- data.frame(name = swiss_names[i], 
                    mean_value = mean_rounder(swiss[, i]))
  means_df <- rbind(tmp, means_df)
  
}
means_df
```

The first line reads: "for each `i` in the vector 1 through the number of columns in the **swiss** data frame . . . "

We then use `i` in the subscripts `[]` to select each column of **swiss** to find and print its mean.

### Apply {#apply-intro}

Loops in R are *notoriously slow*. The `apply` family of functions does a something similar and is usually faster.

For example, let's find the mean of each column in **swiss**:

```{r mean-print}
means_apply <- apply(swiss, MARGIN = 2, mean_rounder)
as.data.frame(means_apply)
```

For data frames and matrices, we need to specify whether we are applying the function to each row (`MARGIN = 1`) or each column (`MARGIN = 2`).

## Debugging in R {#debugging-intro}

Debugging is the process of finding and resolving discrepancies between a program's specification and its implementation.[^7] For example, I want a ggplot2 to plot multiple time series on the same plot as separate lines, but it plots them all as one--very jagged and confusing line. I could use a debugging process to find the cause of and correct this discrepancy.

[^7]: See the [Wiki page on debugging](https://en.wikipedia.org/wiki/Debugging) for a visual example of an early computer bug, an actual bug that got stuck in a computer.

The reason we are covering debugging now, is that statistical simulations are more complex computational and coding processes than we've seen before, so debugging them is more complicated.

### Being mindful of your feelings

We'll cover the the technical debugging tools available in R. But I first want to first mention a crucial part of effective (and less stressful) debugging: *being mindful of your feelings about the bug*.

A bug is frustrating. Our first encounter with a bug is an experience of disassociation between how we think the world should work and how it is working. It's as if we want to put a plate in the cupboard, but each time we try to put the plate in the cupboard it flies out. Frustration inclines us to look for the cause of a bug in--almost always--the wrong place: "the stupid computer is broken".

The best way to resolve bugs, as we'll see, is to create testable hypotheses about what could be causing the bug. "The stupid computer is broken" is vague and so not a testable hypothesis.

So, it is crucial that when you find a bug, to take a step back and ask:

> how does this bug make me feel?
If the answer is "angry, I want to throw my computer out the window", your best next step is to *take a break*. Grab a tea or whatever. Then, with a piece of paper (or some other tool that gets you away from the temptation to just begin slamming out code on your keyboard) write down hypotheses about the cause of the bug and ways to test them.

Unlike social phenomena, most computer bugs are deterministic. The cause of a bug will usually cause the bug. So your chances of definitively finding the bug are high with the following process and tools.

### Clear and testable program specifications

> Finding your bug is a process of confirming the many things that you believe are true --- until you find one which is not true.
>
> ---Norm Matloff (quote from [Advanced R](https://adv-r.hadley.nz/debugging.html#debugging-strategy))

Great debugging is driven by having great expectations of what the program should return. Having a well defined expectation of what a program should do is important to knowing if you have a bug to begin with. One way to approach software development is to write tests for a function *before* you write the function. This is called [test drive development](https://en.wikipedia.org/wiki/Test-driven_development).

Both in your functions, to test incremental expectations, and for the output of your functions you include automated tests. Here is a silly example. Imagine that we have a function that should add 2 + 2 and then divide by 2 to return 2. We can create an automated test like this:

```{r}
# Function to test
two <- function() {
  four <- 2 + 2
  if (four != 4) 
    stop("Did not add 2 + 2 correctly")
  
  four / 2
}
# Automated test 
if (two() != 2) stop("function two did not return two")
```

The `if` . . . `stop` statements automatically test whether or not the function is executing correctly. If there is a problem, the tests should help us pinpoint the cause more quickly.

Automated tests will also help us make sure that our code is correct even as we change it, for example, to add new features.

### `browser()` for testing and dissecting

A key debugging R tool is the `browser()` function. When R executes some code and hits `browser()` it stops the execution and begins an interactive `Browse` mode. This allows you to explore objects created in the code execution in the environment that created then, as well as run new code. RStudio works really well with this function.

Imagine we wrote the loop above incorrectly:

```{r eval=FALSE}
one_sim_bug <- function(n, control_prob, rel_effect) {
  treat_prob <- control_prob / (control_prob * rel_effect)
    
  cy <- rbinom(n = n, size = 1, prob = control_prob)
  ty <- rbinom(n = n, size = 1, prob = treat_prob)
  
  mean(ty) - mean(cy)
}
mean_diff_bug <- rep(NA, n_sims)
for (i in seq_along(mean_diff)) {
  mean_diff_bug[u] <- one_sim_bug(n = n_obs, 
                                  control_prob = control_prob, 
                                  rel_effect = rel_effect)
}
```

We have two layers of nested functions where the bug could be. Let's systematically use `browser()` to identify the source. Start from the outer most function--the `for` loop and work you way down as necessary.

```{r eval=FALSE}
for (i in seq_along(mean_diff)) {
browser()
  mean_diff_bug[u] <- one_sim_bug(n = n_obs, 
                                  control_prob = control_prob, 
                                  rel_effect = rel_effect)
}
```

Ok, we've found the typo. We should have used `i` instead of `u` to index the output of `one_sim`. To leave the browser environment type `c + Enter`. Are we done? We'll let's check the output:

```{r eval=FALSE}
for (i in seq_along(mean_diff)) {
  mean_diff_bug[i] <- one_sim_bug(n = n_obs, 
                                  control_prob = control_prob, 
                                  rel_effect = rel_effect)
}
hist(mean_diff_bug)
```

Hm, we are not getting a valid histogram. Let's now go down the function nesting to insert a `browser()` call in `one_sim`:

```{r eval=FALSE}
one_sim_bug <- function(n, control_prob, rel_effect) {
  browser()
  treat_prob <- control_prob / (control_prob * rel_effect)
    
  cy <- rbinom(n = n, size = 1, prob = control_prob)
  ty <- rbinom(n = n, size = 1, prob = treat_prob)
  
  mean(ty) - mean(cy)
}
mean_diff_bug <- rep(NA, n_sims)
for (i in seq_along(mean_diff)) {
  mean_diff_bug[u] <- one_sim_bug(n = n_obs, 
                                  control_prob = control_prob, 
                                  rel_effect = rel_effect)
}
```

When we do this, we can find that we incorrectly divide `control_prob` by `(control_prob * rel_effect)` the ultimate result, is that the output of `one_sim` is `NA`. We can't create a histogram of `NA`s. Once we solve this issue, we should get the expected histogram.


## Advanced: Parallelisation {#parallel-intro}

By default, all computation in R is done sequentially--one after the other. However, your computer likely has more processor cores and so can run computations in "parallel".

To find the number of cores on your computer use:

```{r detectcores}
library(parallel)
detectCores()
```

Not all computation can be done in parallel. Somethings have to be done in sequence, because the output of one operation is needed for the computation of the subsequent operations. However, pretty much anything that can be looped or applied can be done in parallel. This will often improve the speed of the computation.[^2]

[^2]: It won't necessarily be faster as there is "overhead" for breaking up the operation into parallel computation. Also, while the computation time can go down, the enegery usage will be the same as in total the same computation will have happened.

`mclapply()` from base R's *parallel* package works like `lapply` (list apply) for. Imagine we have a list with 3 numeric vectors:

```{r sim-list}
three_list <- list(A = rnorm(n = 100, mean = 1),
                   B = rnorm(n = 100, mean = 2),
                   C = rnorm(n = 100, mean = 5))
```

Note `rnorm()` simulates data from the normal distribution.

Let's compute the mean of each vector in the list in parallel:

```{r parallel-mean}
mclapply(three_list, mean, mc.cores = 3)
```

`mc.cores` sets the number of cores to use. Usually use 1 - the number of cores you have available.

## 🥅 Exercises

-   Create a new function to summarise numeric vectors in a method of your choosing (explore R's built in statistical capabilities).

-   Create a data frame with numeric and character columns. Then, for each column find the standard deviation (`sd()`) if it is numeric and return the `message()` `"This column has characters"` if it is a character type column.

```{r}
# Create data frame with numeric and character columns
means_df

for (i in seq_along(means_df)) {
  # extract column
  tmp <- means_df[, i]
  if (is.numeric(tmp)) {
    message(sd(tmp))
  }
  else if (is.character(tmp)) {
    message("This is a character vector")
  }
  else
    warning("Not Character or Numeric")
}


```

-   `nchar()` counts the number of characters in a string. Create a list of character vectors with strings of different lengths. In parallel, count the number of characters in each vector.

```{r}


```

## Additional Materials

## (Some) Logical Operators {#logical-operators}

| Logical Operator | Meaning                  |
|------------------|--------------------------|
| `&`              | and                      |
| `|` or           | or                       |
| `==`             | equality                 |
| `<`              | less than                |
| `>`              | greater than             |
| `>=`             | greater than or equal to |
| `<=`             | less than or equal to    |
| `!=`             | not equal                |
| `isTRUE(x)`      | test if `x` is `TRUE`    |

## 📚 References
